Monitoring and Projecting Global Hunger: Are We On Track?

### 2020-12-22 docs/modeling_fies.tex
update title
Changes from Overleaf

\title{Monitoring and Projecting Global Hunger}

@@

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).  This is the first global picture of food security at a subnational level, based on the Food Insecurity Experience Scale, a food security metric developed by FAO that is most indicative of the individual or household lived experience of food insecurity and hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, and global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.  \end{abstract}

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate and severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population. These do not directly capture the actual lived exposure of individuals to food shortages.

The FIES is based on responses to polls and the data are limited to 75 countries where these have been conducted and microdata has been vetted and released by the FAO. The current practice is to assign the average regional prevalence rate for countries without survey data in order to compile regional and global aggregates. In addition, since the FIES is a relatively new metric, it does not have a sufficiently long time series to assess time trends with a reasonable degree of precision. To understand whether SDG 2 is being met, estimates of food insecurity outcomes in 2030 are required.

This paper contributes to shifting the research frontier by filling these two gaps in terms of data coverage and forecasts by developing a statistical model of FIES outcomes based on covariates that are available for territories where actual polling data are not available. Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond the country level to a sub-national level. To our knowledge, this is the first effort to model the FIES outcomes at that scale.

Food security has traditionally been difficult to measure, and this has led to an incomplete or inaccurate assessment of global hunger. Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.  However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.  Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that are food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.  Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.  Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.

As researchers began to focus on food insecurity at the individual and household level, household microdata collecting information on household finances and consumption became a common proxy for food security \citep{Haddad1994}.  However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \citep{Maxwell1996}.  This led to the emergence of several indicators designed to be rapidly deployable, and based on the lived experience of food security \citep{Jones2013}.  These metrics include the Household Food Security Survey Module (HFSSM) \citep{kennedy2005keynote}, originally developed for use in the US; the Latin American and Caribbean Food Security Scale (ELCSA); and the Household Food Insecurity and Access Scale (HFIAS) \citep{Coates2007}.

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Using a Rasch model \citep{Cafiero2018}, each individual in the survey is given a severity score that maximizes the likelihood of observing the responses given to the 8 questions asked in the survey.  Once individuals are ranked in this fashion, thresholds for moderate and severe food insecurity are set that describe people's experiences, and the prevalence of such food insecurity events can be computed. A moderately food insecure adult is likely to have compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure adult is likely to have run out of food and has gone an entire day without eating at times during the year.  Because the model estimates the population over a threshold on a univariate food insecurity scale, the percentage of people over the threshold for moderate food insecurity includes people that are also over the threshold for severe food insecurity.

Since 2014, in partnership with Gallup World Poll, the FAO administered the FIES in surveys around the world. The FAO uses this data to estimate the prevalence of food insecurity, and reports national-level estimates of the percentage of the population over the thresholds for moderate and severe food insecurity.  Drawing on individual-level data from these microdatasets in 75 countries, we use machine learning methods to estimate levels of food insecurity globally at the subnational level.  The model is then used to forecast food insecurity to the year 2030 under a middle-of-the-road scenario.

@@

While economic conditions, and especially GDP per capita, are important predictors of food insecurity at a global scale, there is an upper limit to the extent to which increasing economic development alone can improve food security outcomes.  High income countries have seen little improvement in food insecurity outcomes over twenty years, and upper-middle income countries are expected to see diminishing gains throughout the 2020s.  Among the high-income countries in our dataset, those with the lowest rates of moderate and severe food insecurity also had lower inequality and poverty.  This suggests that our model's forecasted marginal increases in food insecurity in high income countries is due to trends of increasing income inequality, and that economic development alone cannot ensure that everyone is well-fed.

@@

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using leave-one-out cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.

@@

>>>>

\title{Monitoring and Projecting Global Hunger: Are We On Track?}

@@

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by FAO through the Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the World Hunger Clock (https://worldhunger.io).  This is the first global picture of food security at a subnational level based on the Food Insecurity Experience Scale, the indicator for the Sustainable Development Goal of "Zero Hunger" that is most indicative of the individual's lived experience of food insecurity and hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population being at least moderately food insecure in some high-income countries to parts of the developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of severe food insecurity are declining steadily yet slowly, with larger observed and forecasted gains in lower and lower-middle income countries, resulting in global decreases in the total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, we predict it will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population still experiencing hunger by 2030.  \end{abstract}

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate and severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work started in the US in the 1990's to develop a metric of food insecurity based on data gathered by interviewing the people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's more commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population, developed on the basis of infrequent and heterogeneous data on food production and consumption. These do not directly capture the actual lived exposure of individuals to food shortages.

Currently, the data used to generate the estimates that are reported at the UN SDG indicators database come primarily from the 75 countries where surveys have been commissioned, vetted, and released by FAO. To compile global and regional aggregates, the current practice at FAO is to use the prevalence rate for the countries that do have FIES data as the regional rate.  Moreover, since the FIES is a relatively new metric, a sufficiently long time series to assess time trends with a reasonable degree of precision is still lacking. Thus, to fill these spatial and temporal data gaps and understand whether SDG 2 is being met, modeled estimates of food insecurity outcomes are thus required.

This paper contributes to shifting the research frontier by filling these two gaps in terms of spatial coverage and forecasts by developing a machine learning model of FIES-based food insecurity rates based on covariates that are available for territories where actual survey data are not available. This model is then employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FAO's FIES surveys to push the analysis beyond the country level to a sub-national level. To our knowledge, this is the first effort to model FIES-based food insecurity at that scale.

Food security has traditionally been difficult to measure, and this has arguably led to an incomplete or possibly inaccurate assessment of global hunger. Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.  However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.  Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that are food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.  Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.  Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.

As researchers began to focus on food insecurity at the individual and household level, microdata from household surveys collecting information on household finances and consumption became a basis to compute a common proxy for food security \citep{Haddad1994}.  However, these efforts were criticized for being onerous, insufficiently comparable, as well as for ignoring subjective and experiential aspects of food security \citep{Maxwell1996}.  This led to the emergence of several indicators designed to be rapidly deployable, and based on the lived experience of food security \citep{Jones2013}.  These metrics include the Household Food Security Survey Module (HFSSM) \citep{kennedy2005keynote}, originally developed for use in the US; the Latin American and Caribbean Food Security Scale (ELCSA); and the Household Food Insecurity and Access Scale (HFIAS) \citep{Coates2007}.

Drawing on the insights derived in designing and implementing these novel food security metrics, the FIES was developed by the Food and Agricultural Organization (FAO) of the United Nations (UN) \cite{Ballard2013} and is now recognized as a rapidly deployable and cross-culturally valid tool for understanding individual and household-level food insecurity \citep{wambogo2018validity, smith2017world}.  The FIES is based on a survey of eight behaviors indicative of food insecurity and hunger over the previous year, such as skipping meals or worrying about having enough to eat.  Processing the responses given to the FIES questions through the Rasch model, each individual in the survey is assigned a score that allows them to be ranked on a scale of food insecurity severity. Once the severity measures are calibrated against a common reference (the global FIES reference scale) \citep{Cafiero2018} and individuals are ranked in this fashion, conventional thresholds for ``moderate'' and ``severe'' food insecurity are set and the prevalence of such food insecurity conditions can be computed. Thresholds are set in such a way that a moderately food insecure respondent is one who likely has compromised on food quality and variety as a result of lack of resources, has been unsure about their ability to obtain food and has skipped meals or run out of food occasionally. A severely food insecure one is likely to have run out of food and has gone an entire day without eating at times during the year.  The percentage of people estimated to be over the threshold set for moderate food insecurity includes people that are also over the threshold for severe food insecurity and represents the SDG indicator 2.1.2.

Since 2014, FAO contracted the Gallup World Poll to administer the FIES survey module in as many countries as possible around the world and uses this data to estimate the prevalence of food insecurity at region and global levels. For a number of those countries for which it has been authorized to release the national-level estimates, FAO also reports the percentage of the national population that is estimated to be over the thresholds for moderate and severe food insecurity and publicly releases the microdata.

Drawing on individual-level data from these publicly available microdatasets from 75 countries, we use machine learning methods to model food insecurity based on covariates and estimate levels of food insecurity globally, at the subnational level.  The model is then used to forecast food insecurity to the year 2030 under a reasonable scenario on the possible evolution of the covariates.

@@

While economic conditions, and especially GDP per capita, are important predictors of food insecurity at a global scale, there is an upper limit to the extent to which increasing economic development alone can improve food security outcomes.  High income countries have seen little improvement in food insecurity outcomes over twenty years, and upper-middle income countries are expected to see diminishing gains as they continue to develop economically throughout the 2020s.  Among the high-income countries in our dataset, those with the lowest rates of moderate and severe food insecurity also had lower inequality and poverty.  This suggests that our model's forecasted marginal increases in food insecurity in high income countries is due to trends of increasing income inequality, and that economic development alone cannot ensure that everyone is well-fed.

@@

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and food insecurity are the same across all locations.  We test the accuracy of these cases by using leave-one-out cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup on behalf of FAO also records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.

@@

\section{Acknowledgements}
This reserach was supported by The International Fund for Agricultural Development (IFAD) grant number 2000002044.  We thank Kristofer Hamel and Wolfgang Fengler for their support in this work.

\section{Author Contributions}
MC, BM, and HK conceptualized the analysis; CC, JCC, and HK provided methodological oversight; MC, BM, and JCL processed the data; MC and BM ran the models and conducted model validation; All authors contributed to writing and reviewing the manuscript.

\section{Competing Interests}
We have no competing interests to disclose.
